---
title: "Welcome"
---

<br>

This webpage is a semi-opinionated guide to using `ggplot2` for data visualization within social science research. This site was designed as a workshop for graduate students in the social sciences, but folks interested in learning data visualization with R may generally find the code vignettes useful. 

Instead of focusing on the basics of the library (which many guides already do really well--see the resources page) before diving into real-world examples, this workshop takes a somewhat different tack by using common social science use cases to motivate code vignettes that then show how to use ggplot in practice.

The key vignettes linked above are focused on use cases where a graphic can be especially useful at conveying an idea to your reader. In other words, these are instances where we have a clear reason to prefer figures to tables. While these use cases are not intended to be exhaustive, they are intended to be somewhat distinct in data and focus from each other.

<br>

## Requirements

If you want to follow along on your own computer, you'll need to have a recent version of R installed along with the a few accompanying software libraries that we'll use to jump start these vignettes. 

First and foremost of these libraries is the `tidyverse` meta-library which contains `ggplot2` and other useful libraries like `dplyr` that will be referenced within the workshop material. We will also use libraries like `cobalt` within Case A about covariate balance, `fixest` within Case B about design-based methods and `sf` within Case C about spatial clustering. 

```{r setup, message=FALSE}
#all cases
library(tidyverse)

#case a
library(cobalt)

#case b
library(fixest)
library(did)
library(broom)

#case c
library(sf)
library(tigris)
library(ggspatial)
library(biscale)
library(cowplot)
library(ggpattern)

#case d
library(snakecase)
library(patchwork)
```

<br>

To get started in an RStudio session, clone the GitHub repository (which includes all of the necessary data) and follow along with the .Rmd files that correspond to the different vignettes.

<br>

## Session info

```{r session-info}
sessionInfo()
```

